---
title: "Noxious Plants"
output:
  pdf_document: default
  word_document: default
header-includes:
 - \usepackage[width=\textwidth]{caption}
 - \usepackage{wrapfig}
 - \usepackage[font=small,labelfont=bf]{caption}
always_allow_html: yes
csl: ../citations/citations/apa.csl
bibliography: ../citations/citations/citations.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(dpi = 300) 
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
knitr::opts_chunk$set(fig.align="left") 
```

\vspace{-1cm}

One of the first acts in the life of a plant, as a seed, is to move. Generally seeds travel only short distances, but less common long distance movements are known to occur naturally (Section X.X), and long distance movement occurs regularly via humans. The long distance movement of seeds by humans, in recent history *i.e.* the last couple thousand years, may result in the seed germinating and the population (if any) which results from it surviving for only a few generations (*waif*) (@nesom2000non). A seed may also produce a population which is able to persist for many generations in the new ecosystem, and even disperse and spread out across their new landscape (*naturalized*); many of these species are readily incorporated into the existing vegetation and do little to alter it (@nesom2000non, @pyvsek2010invasive). A subset of these species which are able to naturalize may displace considerable amounts of the plant species already in the landscape, and in so doing alters the composition of species at ecological sites (*noxious*), hence also the functioning of these sites (*invasive*) (@davies2011plant, @evans2001exotic, @pyvsek2012global, @noxious2023blm,  *and reviewed in* @ehrenfeld2010ecosystem, @pyvsek2010invasive).  That a few species can alter the properties of landscapes is at the core of the Ecological Sites. While all invasive plants are noxious, not all noxious plants are invasive, as can be seen from a number of native species which act as aggressive weeds in disturbed settings and which prevent natural succession.

Noxious and invasive species adversely affect nearly all ecosystem services offered by natural areas, increase fire frequency and intensity, and have enormous economic impacts (*reviwed in* @ehrenfeld2010ecosystem, @d1992biological, @duncan2004assessing, @fantle2022economic, @crystal2021economic). Invasive species at landscape scales have been shown to decrease plant species richness, taxonomic, functional and structural diversity leading to declines in habitat heterogeneity and adversely affecting wildlife (@keeley2012fire, @ehrenfeld2010ecosystem, @klinger2017alternative). They pose serious threats to the well being of both wildlife, livestock, and humans, via interactions with historic land management alterations, to the fire cycle (@d1992biological, @keeley2012fire, @). The economic impacts of invasive species include enormous amounts of funds being channeled into their treatment to reduce fuel loads (), treatments to curtail their spread into new areas (), and losses in economic activity e.g. by displacement of grasses more suitable as forage for livestock (). 

Most land managers in the Western cold deserts (The Colorado Plateau, Great Basin, and Columbia Plateau) consider noxious annual grasses to pose the greatest challenge towards maintaining ecosystems which allow for multiple uses (@chambers2009cold). Of concern to our office is the increasing adpation of cheatgrass (*Bromus tectorum*), which is already present throughout the field office, towards higher elevation sites (@smith2022elevational). Based on the limited evidence currently available the effect of encroachment of invasive species is of more adverse affect than is attributable to climate change or drought, although synergistic interactions between invasive species and drought still occurr  (@clarke2005long, @lopez2022global). 

# Methods

```{r load libraries}
library(sf)
library(spsurvey)
library(tidyverse)
library(waffle)
library(hrbrthemes)
#install.packages('FSA')

source('functions.R')
p <- file.path('..', 'data', 'raw')
f <- list.files(p)
```

Creation and maintenance of registries of noxious and invasive species often falls on the Department of Agriculture of the Federal and State governments (@quinn2013navigating). Given the focus of these agencies, these lists are generally focused on arable lands used for crop production, with minimal effort placed on natural settings (@quinn2013navigating). To better characterize the condition of invasive and noxious species in natural areas, BLM staff co-developed definitive lists of status of species regarding their Noxious or Invasive status in 2019. 

```{r download records from BIEN, eval = F}

fp <- file.path('..', '..',  'UFO_cartography/BLM_CO_ADMU')
ufo_bbox <- st_read(
  file.path(fp, list.files(fp, pattern = '.shp$')), quiet = T) %>% 
  st_transform(4326) %>% 
  st_bbox()

bien_recs <- BIEN::BIEN_occurrence_box(min.lat  = ufo_bbox[2],
                                 max.lat  = ufo_bbox[4],
                                 min.long = ufo_bbox[1],
                                 max.long = ufo_bbox[3])
write.csv(bien_recs, file.path('..', 'data', 'raw', 'bien_occurrence_from_bbox.csv'))

rm(fp, ufo_bbox, bien_recs)
```

```{r identify invasive species from CNHP, eval = F}

attributes <- read.csv( file.path(p, f[grep('Attributes', f)])) %>% 
  distinct(Species, Noxious ) %>% 
  filter(Noxious == 'YES')

invasive_species <- read.csv(file.path(p, f[grep('Values', f)])) %>% 
  select(National_USDASymbol, National_SciName_noAuthority, Ack_SciName_noAuthority,  National_NativeStatus, 
         CONox_Status, Invasiveness, IRank, IRank_Score) %>% 
  mutate(across(.col = everything(), ~ na_if(., ''))) %>% 
  drop_na(National_USDASymbol) %>% 
  filter(National_NativeStatus %in% c('Introduced', 'Both')) %>% 
  unite(CONox_Status:IRank_Score, col = 'Invasive' , remove = F, na.rm = T) %>% 
  mutate(Invasive = na_if(Invasive, "")) %>% 
  mutate(Invasive = if_else(!is.na(Invasive), T, F))

attributes[!attributes$Species %in% invasive_species$National_USDASymbol,]
# these mostly due to the CPNWH being finer in taxonomy, e.g. if only an infraspecific 
# taxon known to the state, than this recorded.

aim_synonyms <-  read.csv( file.path(p, f[grep('AIM_invasive', f)])) 

invasive_species %>% 
  mutate(National_USDASymbol = if_else(
    National_USDASymbol == 'TACH$', 'TACH2', National_USDASymbol
  )) 
invasive_species <- bind_rows(aim_synonyms, invasive_species)

write.csv(invasive_species, row.names = F, 
          file.path('..', 'data', 'processed', 'Introducted_species_CO.csv'))

rm(attributes, invasive_species)
```

To develop a list of invasive plant species for the study area 'IRanks', a semi-quantitative expert based assessment of introduced species, were extracted from the C-Values prepared by the Colorado Natural Heritage Program (Section X.X) (@moore2004iranks, @cnhp2020fqi).  This was combined with an AIM species attribute table extract from the vicinity of the study area, the latter dataset contained a handful of synonyms which included codes not present in the former.

Once this list was developed and underwent review, the cleaned AIM data maintained on TerraDat had both the Species Richness and Line-point Intercept data downloaded, and reprocessed to determine both the presence and absence of species and, when present at LPI location their percent fractional cover. In order to determine what percent of the field office were meeting benchmark reference condition, which is the same for all Ecological Sites - the lack of invasive species, if an invasive was detected on a plot it was considered to be failing to meet benchmarks. These plots then underwent categorical analysis using *cat_analysis*, in the 'spsurvey' package, with confidence interval of 0.8, (@dumelle2022sp), and the 'local' (default) variance estimator.

```{r import species richness data and tie to plots}

plot_char <- read.csv( file.path(p, f[grep('Characterization', f)]) ) %>% 
  select(PrimaryKey, Easting, Northing) %>% 
  drop_na() 

invasive_species <- read.csv(
  file.path('..', 'data', 'processed', 'Introducted_species_CO.csv') ) %>% 
  select(National_USDASymbol:Invasive)

spp_alpha <- read.csv( file.path(p, f[grep('Richness.csv$', f)])) %>% 
  select(PrimaryKey, SpeciesList, abundance, distribution) %>% 
  right_join(., plot_char, by = 'PrimaryKey') %>% 
  st_as_sf(coords = c('Easting', 'Northing'), crs = 4269) %>% 
  st_transform(26913)

pts <- st_read(
  '/media/sagesteppe/ExternalHD/aimDB/data/raw/AIM_Sample_Design/AIM_Design_Stratification.shp',
                quiet = T) %>% 
  st_transform(26913) %>% 
  st_buffer(100) %>% 
  select(PLOTID, STRATUM)

spp_alpha <- spp_alpha %>% 
  st_transform(26913)

esd_class <- read.csv( file.path(p, f[grep('Tracking', f)]) ) %>% 
  filter(STATUS == 'SAMPLED') %>% 
  mutate(ECO.SITE = if_else(ECO.SITE.MATCHED == F, 'UNKNOWN', ECO.SITE)) %>% 
  select(PlotID = PLOT.ID, ECO.SITE)

spp_alpha <- st_intersection(pts, spp_alpha) %>% 
  left_join(., esd_class, by = c('PLOTID' = 'PlotID')) %>% 
  mutate(ECO.SITE = if_else(PLOTID == 'GR-035', 'R036XY114CO', ECO.SITE),
         PLOTID = if_else(PLOTID == 'GR-035', 'GR-025', PLOTID)) 

rm(plot_char, pts, esd_class)
```

```{r import LPI data and tie to plots}

lpi <- read.csv( file.path(p, f[grep('LPIRAW', f)]) ) %>% 
  select(PrimaryKey, TopCanopy:SoilSurface) %>% 
  filter(PrimaryKey %in% spp_alpha$PrimaryKey) %>% 
  mutate(TopCanopy = na_if(TopCanopy, 'None'),
               TopCanopy = na_if(TopCanopy, 'N')) %>% 
  mutate(across(.cols = everything(), ~ na_if(.x, ""))) %>% 
  unite(col = 'All_species',  TopCanopy:SoilSurface,  sep = " ", na.rm = T, remove = F) 

Invasive <- lapply(lpi$All_species, inv_present, invasive_species$National_USDASymbol)
Invasive <- unlist(Invasive)
lpi <- bind_cols(lpi, Invasive = Invasive)

# percent cover of all plants
lpi_sum <- lpi %>% 
  group_by(PrimaryKey) %>% 
  dplyr::summarize(Vascular_Intercepts = 150 - sum(is.na(TopCanopy)),
            Invasive_Intercepts = sum(Invasive))

# join up
lpi_sum <- left_join(lpi_sum, select(spp_alpha, PLOTID, PrimaryKey) %>% 
                       distinct(), by = 'PrimaryKey') %>% 
  select(-PrimaryKey) %>% 
  mutate(Proportion_Invasive = Invasive_Intercepts/Vascular_Intercepts) %>% 
  st_as_sf()

# Cover of individual invasive species by plots

lpi2 <- lpi %>% 
  select(PrimaryKey:Lower4, -All_species) %>% 
  pivot_longer(!PrimaryKey, values_to = 'SYMBOL')  %>% 
  filter(!SYMBOL %in% c('L', 'WL', 'HL')) %>% 
  drop_na(SYMBOL) %>% 
  group_by(PrimaryKey, SYMBOL) %>% 
  count(SYMBOL) %>% 
  filter(SYMBOL %in% invasive_species$National_USDASymbol) %>% 
  ungroup()

invasive_sum <- left_join(lpi2, select(spp_alpha, PLOTID, PrimaryKey) %>% 
                       distinct(), by = 'PrimaryKey') %>% 
  select(-PrimaryKey, -geometry) %>% 
  left_join(., select( invasive_species, National_USDASymbol, Invasive), 
            by = c('SYMBOL' ='National_USDASymbol'))

rm(Invasive, inv_present, lpi2)
```

```{r Join invasive species and process raw data}

inv <- spp_alpha %>% 
  left_join(., invasive_species, by = c('SpeciesList' = 'National_USDASymbol')) %>% 
  group_by(PLOTID) %>% 
  add_count(name = 'No_Spp') %>% 
  group_by(PLOTID) %>% 
  mutate(INTRO_CNT = if_else(National_NativeStatus %in% c('Both', 'Introduced'), 1, 0 ),
         No_Introduced = sum(INTRO_CNT)) %>% 
  mutate(INVAS_CNT = if_else(Invasive == T, 1, 0 ),
         No_Invasive = sum(INVAS_CNT, na.rm =T)) %>% 
  select(-INVAS_CNT, -INTRO_CNT) %>% 
  ungroup(PLOTID)

inv_richness <- inv %>% 
  drop_na(National_SciName_noAuthority) %>% 
  mutate(Prop_Int = No_Introduced/No_Spp,
           Prop_Inv = No_Invasive/No_Spp)

inv_plots <- select(inv, PLOTID, ECO.SITE, No_Spp, No_Introduced, No_Invasive)  %>% 
  group_by(PLOTID) %>% 
  slice_max(No_Invasive, n = 1, with_ties = F)

rm(spp_alpha, inv)
```

```{r Rescale three Proportions from zero to one and sum together for cartography, fig.height=3.5, fig.width =3.5}

inv_carto <- inv_richness %>% 
  mutate(across(.cols = Prop_Int:Prop_Inv, ~ scales::rescale(., to = c(0,1)))) %>% 
  select(PLOTID, Prop_Int, Prop_Inv) %>% 
  st_drop_geometry()

inv_carto <- left_join(lpi_sum, inv_carto, by = "PLOTID") %>% 
  distinct(PLOTID, .keep_all = T) %>% 
  rowwise() %>% 
  mutate(Prop_Inv2 = sum(Proportion_Invasive + Prop_Inv + Prop_Int, na.rm = T)/3) %>% 
  select(PLOTID, Index_Prop = Prop_Inv2, Proportion_Invasive, Prop_Inv, Prop_Int)

plot_vars <- inv_carto %>% 
  st_drop_geometry() %>% 
  pivot_longer(cols = c('Prop_Inv', 'Prop_Int', 'Proportion_Invasive'), 
               names_to = 'Variable', values_to = 'Proportion') %>% 
  mutate(Proportion = replace_na(Proportion, 0))

cols <- c("Prop_Int" = '#f79489', 'Prop_Inv' = '#F6131E',
          'Proportion_Invasive' =  '#9D060D')
labs <- c('(Spp) Introduced', '(Spp) Invasive', '(Cover) Invasive')

#summary( glm(Index_Prop ~  Proportion + Variable, family = "poisson", data = plot_vars) )

ggplot(data = plot_vars, aes(x = Index_Prop, y = Proportion)) +
  geom_point(aes(color = Variable, shape = Variable)) +
  geom_smooth(colour = '#391306', method = 'glm') +
  scale_color_manual('Proportion', values = cols, 
                     labels = labs) +
  scale_shape_manual('Proportion', values = c(0, 1, 2), 
                     labels = labs) +
  theme_bw(base_size = 10) +
  theme(aspect.ratio = 1, legend.title = element_text(hjust = 0.5)) +
  labs(title = 'Composite "Invasive Index"', x = 'Invasive Index', 
       y = 'Proportion of Variables') 

```

```{r Write out data for three index plot, eval = F}

carto_path <- '../../UFO_cartography'
if (!dir.exists(file.path(carto_path, 'noxious'))) {dir.create(file.path(carto_path, 'noxious'))}
st_write(inv_carto,  file.path(carto_path, 'noxious', 'noxious_index.shp'))

```

```{r Perform Weighted Analysis of Noxious Plants, warning = F}

wghts <- read.csv( file.path(p, f[grep('Weights', f)])) %>% 
  select(WghtPerPlot, STRATUM = Stratum, AoInference = Area)

esd_class <- read.csv( file.path(p, f[grep('Tracking', f)]) ) %>% 
  filter(STATUS == 'SAMPLED') %>% 
  mutate(ECO.SITE = if_else(ECO.SITE.MATCHED == F, 'UNKNOWN', ECO.SITE)) %>% 
  select(PLOTID = PLOT.ID, ECO.SITE)

# Perform with WHETHER PLANT was ON PLOT (species richness)
inv_plots <- inv_plots %>%  
  mutate(Meeting = if_else(No_Invasive == 0, T, F)) 

inv_plots <- read.csv( file.path(p, f[grep('Interence', f)])) %>% 
  select(PLOTID = Plot.ID, AoInference) %>% 
  right_join(., inv_plots) %>% 
  mutate(STRATUM = str_remove(str_extract(PLOTID, '^.*-'), '-')) %>% 
  left_join(., wghts, by = c('STRATUM','AoInference')) %>% 
  left_join(., esd_class, by = 'PLOTID') %>% 
  select(-ECO.SITE.y, ECO.SITE = ECO.SITE.x) %>% 
  st_as_sf()

noxious_presence_analysis <- cat_analysis(
  dframe = inv_plots, siteID = 'PLOTID', vars = 'Meeting',
  subpops = 'AoInference',  weight = 'WghtPerPlot', vartype = 'Local', conf = 80)

# Perform with WHETHER PLANT was ON LPI line 
meeting <- lpi_sum %>% 
  mutate(Meeting = if_else(Proportion_Invasive  == 0, T, F)) %>% 
  select(PLOTID, Meeting, Proportion_Invasive)

meeting <- read.csv( file.path(p, f[grep('Interence', f)])) %>% 
  select(PLOTID = Plot.ID, AoInference) %>% 
  right_join(., meeting) %>% 
  mutate(STRATUM = str_remove(str_extract(PLOTID, '^.*-'), '-')) %>% 
  left_join(., wghts, by = c('STRATUM','AoInference')) %>% 
  left_join(., esd_class, by = 'PLOTID') %>% 
  st_as_sf()

noxious_cover_analysis <- cat_analysis(
  dframe = meeting, siteID = 'PLOTID', vars = 'Meeting',
  subpops = 'AoInference',  weight = 'WghtPerPlot', vartype = 'Local', conf = 80)

rm(wghts, esd_class, warn_df, inv_carto, f, p)
```

```{r Do Certain Ecosites appear less resistant to invasion}

m1 <- meeting %>% 
  group_by(ECO.SITE) %>% 
  filter(ECO.SITE != 'UNKNOWN', n() >= 3) 

kt <- kruskal.test(m1$Proportion_Invasive, as.factor(m1$ECO.SITE))
dt_res <- FSA::dunnTest(Proportion_Invasive ~ as.factor(ECO.SITE),
  data = m1,
  method = "holm"
)[['res']] %>% 
  filter(P.adj < 0.8)

Mdn <- group_by(m1, ECO.SITE) %>% 
  dplyr::summarize(count = n()) %>% 
  ungroup() %>% st_drop_geometry() 

rm(dt_res)
```

To detect whether Ecological Sites differed in their resistance to noxious weed invasion a Kruskall-Wallis test was used (@kruskal1952use). Kruskal-Wallis was used due to non-normal data (heavily right skewed), and a small number of replicates for each Ecological Site. The Kruskall-Wallis test offered evidence of a difference (p = `r round(kt[["p.value"]], 3)`), to detect which Ecological Sites differed from others a two-sided Dunn's test with Holms correction for multiple testing was used as a *post-hoc* test (@dinno2017dunn, @ogle2022fsa, @holm1979simple, @dunn1964multiple). Once p-values were adjusted for multiple testing, as we had no *a-priori* hypothesis for which sites would be more resistant, there was no strong statistical evidence that any pairs were significantly different than any others. However, this is most certainly in part due to the very few number of plots per sites (Mdn = `r median(Mdn$count)`), and the large number of sites (total = `r length(unique(m1$ECO.SITE))`), rather than an actual lack of resistance in Ecological Sites.

```{r remove some writing variables}
rm(kt, m1, Mdn, meeting)
```

```{r Gather localities of weeds for mapping}

r <- left_join(select(lpi_sum, PLOTID, Vascular_Intercepts) %>% distinct(), 
               invasive_sum, by= 'PLOTID')  %>% 
  rowwise() %>% 
  mutate(Prop_Cover = n/Vascular_Intercepts) %>% 
  filter(Prop_Cover <= 1.0) # need to see why SD-338
  
r_locations <- r %>% 
  group_by(SYMBOL) %>% 
  add_count(SYMBOL) %>% 
  filter(Invasive == T)

ne_weeds <- c('ALDE','CACH42','CADR','CIAR4', 'COAR4', 
                'CYOF', 'GYEL', 'LEDE', 'PHPR3', 'POAR8', 'POAV', 'TARA')
ne_labels <- filter(invasive_species, National_USDASymbol %in% ne_weeds) %>% 
  pull(National_SciName_noAuthority)

r_locations_ne <- r_locations %>% filter(SYMBOL %in% ne_weeds)
r_locations <- r_locations %>% filter(!SYMBOL %in% ne_weeds)

rm(lpi, vasc_total, r, invasive_species, ne_weeds, ne_labels)
```


```{r  write out data for mapping in UFO elements of style, eval = F}

st_write(r_locations_ne, file.path(carto_path, 'noxious', 'noxious_NE.shp'), 
         quiet = T)
st_write(r_locations, file.path(carto_path, 'noxious', 'noxious_all.shp'), 
         quiet = T, append = F)

```

# Results

```{r Prepare some writing results}

inv_richness <- st_drop_geometry(inv_richness)
inv_r_t <- inv_richness[inv_richness$Invasive==T,]

most_common_inv <-  sort(
      table(inv_r_t$National_SciName_noAuthority))[
        length(unique(inv_r_t$SpeciesList)):(length(unique(inv_r_t$SpeciesList))-5)]

rm(r_locations_ne, r_locations)
```


\begin{wrapfigure}{r}{0.6\textwidth}
  \centering
    \includegraphics[width=0.6\textwidth]{../results/maps/allFO_weeds.png}
  \caption{Distribution of Invasive Species Across the Field Office}
\end{wrapfigure}

`r length(unique(inv_richness$SpeciesList))` introduced, and of these `r length(unique(inv_r_t$SpeciesList))` invasive species were detected across all `r` plots. `r length(unique(inv_r_t$PLOTID))` plots had invasive species on them, and `r group_by(inv_r_t, PLOTID) |> filter(n() >=2) |> distinct(PLOTID) |> nrow()` plots had more than one invasive species. When reducing these to species of immediate management concern `r` species were present at `r` plots (REVIEW THE LIST AND THIN IT). As expected, the species which occurred the most often were `r toString(names(most_common_inv))` at `r toString(as.numeric(most_common_inv))` of all plots respectively. Most of the species were present across the entirety of the field office, except for ca. a dozen species which were isolated to localities between Delta and Paonia (Figure XX). Several of these species generally occur in wetter, higher elevation, habitats than most terrestrial BLM Land and their spread to BLM is minimal; however a number of the populations are adjacent to USFS land (Figure XX). 

A component of reference condition for all Ecological Sites is that invasive species are not present. By this metric the plots which would be meeting their benchmarks is quite low (`r length(unique(inv_r_t$PLOTID))`), however we feel another consideration is whether an invasive species was detected on the Line-Point Intercepts (Section XX). As reference condition benchmarks still have considerable association with pre-Columbian times ecology, and much has changed in the interim, we believe each of these plots deserves another more liberal consideration. Of the plots which had invasive species, they were detected via line-point intercept at `r  nrow(lpi_sum[lpi_sum$Proportion_Invasive>0,])` of them.  

\begin{wrapfigure}{l}{0.7\textwidth}
  \centering
    \includegraphics[width=0.7\textwidth]{../results/maps/invasive_index.png}
  \caption{Invasive Index across the Field Office}
\end{wrapfigure}

There were `r nrow(unique(invasive_sum[invasive_sum$Invasive == T,'SYMBOL'] ))` species which were detected on LPI lines at `r nrow(lpi_sum[lpi_sum$Proportion_Invasive>0,])` plots, and of these species `r` were of immediate management concern.  The proportion of all vegetation cover which was invasive at plots varied from `r min(lpi_sum$Proportion_Invasive)`% to `r max(lpi_sum$Proportion_Invasive) * 100`% (Mdn = `r round(median(lpi_sum$Proportion_Invasive) * 100, 2)`, mean = `r round(mean(lpi_sum$Proportion_Invasive) * 100,2)`). In general, the few plots with exceptionally high cover of invasive species were small parcels of BLM land adjacent to roads, and private lands, in low elevation areas near Delta (figure XX). 

A caveat with detecting invasive species is that, in the study area, most of them are annuals (##/##). Given the exceptional drought conditions (SECTION XX) under which field work was conducted, it is quite likely that the estimated covers of them by plot are notably lower than are during more normal conditions (@mack1984demography, @bowers1987precipitation). While it seems that their presence at plots is unlikely to be altered, their abundances were likely underestimated in most years of sampling (cite). The relationships between the proportion of introduced species as a function of all species at a plot, the proportion of invasive species (a subset of introduced species), and the proportion of invasive species cover at a plot were highly correlated, and they were combined into a single metric, 'Invasibility Index' (Figure 1). We suspect that this index is more indicative of the status of a plot in a year with more normal precipitation, as can be seen in Figure 1, the data from the time period generally has a large cluster of invasive cover in the lower left of the plot from (0 - 0.25 on the index axis, beneath the line), which is less representative of the potential for invasive cover to expand rapidly under normal conditions. 


\begin{wrapfigure}{r}{0.6\textwidth}
  \centering
    \includegraphics[width=0.6\textwidth]{../results/maps/NothernFO_weeds.png}
  \caption{Uncommon Invasives in the Northern Field Office}
\end{wrapfigure}

A handful of the invasive species detected (Whitetops (*Cardaria chalepensis*, *Cardaria draba*, alternatively *Lepidum draba*), Canada thistle (*Cirsium arvense*), field bindweed (*Convolvulus arvensis*),  Timothy (*Phleum pratense*)), especially in the Northern Portion of the field office are generally associated with slightly more mesic conditions than occur at most BLM land, especially adjacent to irrigated pastures. Unless these populations are entrenched near streams, their spread is likely curtailed by the general aridity of BLM land, but their successful extirpation, in the face of continual re-colonization from adjacent pasture lands, or if they have invaded wetlands, is unlikely. Baby's Breath (*Gypsophila elegans*) may have been introduced for roadside plantings, and is worth eradication efforts (@pringle1993fna). Houndstongue *Cynoglossum officinale* is generally limited to higher elevation forested areas, and in the absence of forest fire it's spread may be slow on the portions of BLM land which it inhabits. A couple of these (Prairie Pepperweed (*Lepidium densiflorum*), prostrate knotweed (*Polygonum aviculare*)) are generally associated with heavily compacted soils, and tend to not spread aggressively outside of these areas. 


```{r Prepare figures of amount of field office meeting objectives}

waff_dat <- noxious_cover_analysis %>% 
  filter(Category == T) %>% 
  mutate(
    across(where(is.numeric), ~ round(.x, 0)),
    Low = if_else(LCB80Pct.P == 0, 0, LCB80Pct.P - 1),
    CI_Low = Estimate.P - LCB80Pct.P,
    Estimate = 1,
    CI_High = UCB80Pct.P - Estimate.P,
    High = if_else(UCB80Pct.P == 0, 0, 100 - UCB80Pct.P)
  ) %>% 
  rowwise(Subpopulation) %>% 
  mutate(High = if_else(sum(c_across(Low:High)) > 100, High -1, High),
         High = if_else(sum(c_across(Low:High)) < 100, High + 1, High))  %>% 
  select(Subpopulation, Low:High) %>% 
  pivot_longer(!Subpopulation, names_to = 'Statistic', values_to = 'Value') %>% 
  group_by(Subpopulation) %>% 
  mutate(
    Statistic = factor(Statistic,
                       levels = c('Low', 'CI_Low', 'Estimate', 'CI_High', 'High')), 
    Subpopulation = factor(Subpopulation,
                       levels = c('ACEC-WSA', 'DE-NM', 'GG-NM', 'SampleFrame')))

waff_dat <- data.table::setorder(waff_dat, Statistic)
waf_values = c('Low' = '#1FE032', 'High' = '#E0321F', 'CI_Low' = '#808929', 
               'Estimate' = 'black', 'CI_High' = '#808929')


ggplot(waff_dat, aes(fill = Statistic, values = Value)) +
  geom_waffle(color = "white", size = .25, flip = TRUE) +
  facet_wrap(~ Subpopulation) +
  scale_fill_manual('Condition', values = waf_values, 
                    labels = c('Meeting', 'Conf. Int.', "Estimate", 'Conf. Int.', 'Failing')) +
  scale_x_discrete()  +
  scale_y_continuous(labels = function(x)round(x * 10, -1), expand = c(0,0))  +
  
  theme(
    legend.position="left",
    aspect.ratio = 1,
    legend.key.size = unit(0.7, 'lines'),
    legend.title = element_text(hjust = 0.5, size = 10), 
    plot.title = element_text(hjust = 0.5, size = 11, face = 'bold'),
    panel.spacing = unit(0.01, "lines"), 
    strip.background =element_rect(fill="white"), 
    panel.background = element_blank())

waff_dat %>% group_by(Subpopulation) %>% mutate(t = sum(Value))

```

A wide range of invasive species were present throughout the field office in a variety of habitats, and at varying abundances (Figure XX). All four areas of analysis within the field office, the ACEC's - WSA's, and the Dominguez-Escalente and Gunnison Gorge National Monuments failed to meet management objectives for being in reference condition (80% of the first three, and 70% of the remaining land) (Figure XX, panel 1). 
They further all failed objectives if invasives were reduced to only presences on LPI positions (Figure XX, panel 2). 

```{r}
rm(most_common_inv, inv_r_t, inv_richness, lpi_sum, invasive_sum)
```



\newpage

# References

